Forecasting the 2024 U.S. Presidential Election

### 2024-10-24 paper/paperV2.qmd
version 2 plan

---
title: "Forecasting the 2024 U.S. Presidential Election"
subtitle: "My subtitle if needed"
author: 
  - Tim Chen
  - Steven Li
  - Tommy Fu
thanks: "Code and data are available at: https://github.com/timchen0326/US_presidential_election_forecast_2024.git"
date: today
date-format: long
abstract: "Analysis of polling data from July to October 2024 reveals significant challenges in forecasting the U.S. presidential election between Kamala Harris and Donald Trump, with state-level variations and pollster methodology playing crucial roles in prediction accuracy. Using multi-linear regression models incorporating pollster ratings, sample sizes, and state-level data, we find that Harris is projected to receive 216 electoral votes while Trump is predicted to secure 147 electoral votes. State-level polling variations and methodological differences among pollsters significantly impact predicted outcomes, with pollster reliability metrics explaining substantial variance in candidate support levels. These findings highlight the continued importance of refining polling methodologies and expanding state-level coverage to improve electoral forecasting accuracy, particularly in an era of increasing political polarization."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

@@

Recent history has highlighted the complexities of election forecasting. The polling failures in 2016 and 2020—where polls significantly underestimated Republican support in key states—have prompted a fundamental reassessment of polling methodologies [@keeter2024]. These challenges are particularly acute in swing states, where margins of victory are often razor-thin and can determine the outcome of the entire election. The American Association for Public Opinion Research (AAPOR) identified several critical factors contributing to these polling errors, including the underrepresentation of Republican voters and difficulties in predicting voter turnout patterns [@viala2024].

@@

This paper develops statistical models to forecast the outcome of the 2024 presidential election between Kamala Harris and Donald Trump. By leveraging multi-linear regression models, we predict the percentage of support for each candidate across different states, incorporating key variables such as pollster rating, sample size, and state-level demographics. Through aggregating these state-level predictions, we simulate Electoral College outcomes to provide insights into each candidate's probability of securing the required 270 electoral votes. Our analysis also includes a detailed examination of YouGov's polling methodology and proposes an idealized survey approach that could enhance the accuracy of election forecasting.

@@

@@

- **Pollster rating (Pollscore):** A numerical score reflecting the reliability and historical accuracy of each polling organization.
- **Sample size:** The number of respondents included in each poll, which influences the poll's margin of error.
- **Support percentage (pct):** The percentage of respondents expressing support for each candidate.
- **State:** The U.S. state where the poll was conducted, or in some cases, national-level polling data.
- ...

@tbl-overview_data presents a sample of the processed dataset, showcasing the key variables necessary for the analysis.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-overview_data
#| tbl-cap: Sample Overview of Selected Variables in the Polling Dataset

kable(data_overview)
```

@@

**Edit note: Add a table in appendix to show sample with all variables used in model**

@@

# Modeling Support for the Candidates {#sec-model}

## Kamala Harris

We begin by modeling the percentage of support for Kamala Harris using a linear regression model.
The predictors include the sample size, pollster ratings (e.g., pollscore and transparency score), and state.
This model aims to quantify how these variables influence her support across different polls.

```{r}
#| echo: false
#| message: false
#| warning: false
# Load necessary library
library(ggplot2)

# Filter the data for only "Kamala Harris"
harris_data <- subset(data, candidate_name == "Kamala Harris")

# Perform a simple linear regression with pct as the dependent variable and sample_size as the independent variable
model <- lm(pct ~ sample_size, data = harris_data)
```

The results of the linear regression for Kamala Harris show that sample size has a statistically insignificant effect on her support.
This suggests that factors other than sample size, such as pollster methodology or regional biases, may play a more significant role in determining the level of support she receives.

@fig-harris-samplesize-supportpct illustrates the relationship between the sample size and support percentage for Kamala Harris.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-harris-samplesize-supportpct
#| fig-cap: Linear Regression of Percentage vs Sample Size for Kamala Harris

# Plot the relationship for Kamala Harris
ggplot(harris_data, aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(x = "Sample Size",
       y = "Percentage")
```

## Donald Trump

A similar linear regression model was applied to Donald Trump's data.
The predictors remain the same, and the goal is to determine the factors that drive support for him.

```{r}
#| echo: false
#| message: false
#| warning: false
# Load necessary library
library(ggplot2)

# Filter the data for only "Kamala Harris"
trump_data <- subset(data, candidate_name == "Donald Trump")

# Perform a simple linear regression with pct as the dependent variable and sample_size as the independent variable
model <- lm(pct ~ sample_size, data = trump_data)

The results in indicate that sample size has a weak but statistically significant negative effect on support for Trump.
This suggests that larger polls tend to show slightly lower support for Trump, although the effect size is small.

@fig-trump-samplesize-supportpct shows the relationship between the sample size and support percentage for Donald Trump.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-trump-samplesize-supportpct
#| fig-cap: Linear Regression of Percentage vs Sample Size for Donald Trump

# Plot the relationship for Trump
ggplot(trump_data, aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(x = "Sample Size",
       y = "Percentage")

mlr_harris_model <- readRDS(here::here(("models/mlr_harris_model.rds")))
mlr_trump_model <- readRDS(here::here(("models/mlr_trump_model.rds")))
```

## Multiple Linear Regression Models

To better capture the complexity of voter support, we constructed multiple linear regression (MLR) models for both Kamala Harris and Donald Trump.
These models incorporate several predictors, including pollster rating (represented by pollscore), transparency score, sample size, and state-level data.
By accounting for these factors, the MLR models allow us to control for more variables that influence voter support and provide more accurate predictions.

### Kamala Harris Model

The MLR model for Kamala Harris takes into account multiple factors that may impact her support across different polls. This includes not just the sample size of the poll, but also how pollster reliability (numeric_grade), transparency of poll data (transparency_score), and state-level polling contribute to predicting voter preferences.

The results @fig-mlr-harris show that several predictors, including state and pollster rating, significantly impact her predicted percentage of voter support.
These results suggest that voter preferences for Harris vary widely depending on the state and the credibility of the pollster.

The model diagnostics (such as residual plots, QQ plots) were evaluated to ensure the assumptions of linear regression hold.
Figure X shows diagnostic plots, which indicate that the model performs reasonably well in terms of residual behavior and normality.
Similarly, we build an MLR model for Donald Trump.

### Donald Trump Model

Similarly, we constructed an MLR model for Donald Trump using the same set of predictors to assess the factors that influence his support across the country.

The results for Donald Trump in @fig-mlr-trump reveal that state-level variations and pollster transparency play a significant role in explaining his level of support.
The model for Trump also passes diagnostic checks, as shown in Figure Y, indicating that the assumptions of linearity, independence, and homoscedasticity are reasonably met.
Check Multi-collinearity using Variance Inflation Factor (VIF)

## Multicollinearity Check Using Variance Inflation Factor (VIF)

To ensure that the predictors used in both models do not exhibit multi-collinearity, we checked the Variance Inflation Factor (VIF) for each predictor. High VIF values indicate multi-collinearity, which can affect the stability and reliability of the model coefficients.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-harris-vif
#| tbl-cap: Harris MLR model Variance Inflation Factor (VIF)

# Check VIF for Kamala Harris model
kable(vif(mlr_harris_model), digits=3)

mlr_harris_model_refined <- readRDS(here::here(("models/mlr_harris_model_refined.rds")))
```

Based on the VIF results in @tbl-harris-vif, we refined the model by removing less significant predictors (such as methodology and transparency_score) to reduce multi-collinearity.This improves the model's accuracy and interpretability.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-harris-refined-vif
#| tbl-cap: Harris Refined MLR model Variance Inflation Factor (VIF)

# Refine the model by removing less significant predictors 
# (e.g., methodology and transparency_score)

# Check VIF for the refined model
kable(vif(mlr_harris_model_refined), digits=3)

step_model <- step(mlr_harris_model_refined, direction = "both", trace=0)
```

@@

To further optimize the MLR models, we performed stepwise model selection, which systematically adds or removes predictors to minimize the Akaike Information Criterion (AIC) and improve model fit.

The stepwise model selection improved the model by retaining the most significant predictors and eliminating those with little explanatory power.

## Final Predictions for Electoral College Votes

Using the final models for both candidates, we predicted the percentage of support in each state and aggregated the results to simulate the Electoral College outcome.
The predicted percentage of support for each candidate is used to determine the likely winner in each state.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ccc
#| fig-cap: ccc

# Final models for Harris and Trump
mlr_harris_model_final <- readRDS(here::here(("models/mlr_harris_model_final.rds")))
mlr_trump_model_final <- readRDS(here::here(("models/mlr_trump_model_final.rds")))

# Predict poll percentages for Kamala Harris
harris_data$predicted_pct_harris <- predict(mlr_harris_model_final, newdata = harris_data)

# Predict poll percentages for Donald Trump
trump_data$predicted_pct_trump <- predict(mlr_trump_model_final, newdata = trump_data)

# Electoral College Prediction {#sec-predict}

To forecast the winner of the 2024 election, we aggregate the predicted percentages of support from our models for each state and calculate the Electoral College votes.
The candidate with 270 or more Electoral College votes is predicted to win the election.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ddd
#| fig-cap: ddd

# Define electoral votes for each state (and Washington, D.C.)
electoral_votes <- data.frame(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida", 
            "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", 
            "Maine CD-1", "Maine CD-2", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
            "Missouri", "Montana", "Nebraska", "Nebraska CD-1", "Nebraska CD-2", "Nebraska CD-3", "Nevada", 
            "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
            "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", 
            "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming", 
            "District of Columbia"),
  electoral_votes = c(9, 3, 11, 6, 55, 9, 7, 3, 29, 16, 4, 4, 20, 11, 6, 6, 8, 8, 2, 1, 1, 10, 11, 15, 10, 6, 10, 
                      3, 5, 1, 1, 2, 6, 4, 14, 5, 29, 16, 3, 18, 7, 6, 20, 4, 9, 3, 11, 38, 6, 3, 13, 12, 5, 10, 3, 3)
)

# Aggregate predictions by state for Harris and Trump
harris_state_avg <- aggregate(predicted_pct_harris ~ state, data = harris_data, FUN = mean)
trump_state_avg <- aggregate(predicted_pct_trump ~ state, data = trump_data, FUN = mean)

# Merge predictions for both candidates
prediction_comparison <- merge(harris_state_avg, trump_state_avg, by = "state", all.x = TRUE)

# Merge the electoral votes with the predictions
prediction_comparison <- merge(prediction_comparison, electoral_votes, by = "state", all.x = TRUE)

# Determine the winner for each state
prediction_comparison$winner <- ifelse(prediction_comparison$predicted_pct_harris > prediction_comparison$predicted_pct_trump, "Harris", "Trump")

# Calculate total electoral votes for Kamala Harris
harris_electoral_votes <- sum(prediction_comparison$electoral_votes[prediction_comparison$winner == "Harris"], na.rm = TRUE)

# Calculate total electoral votes for Donald Trump
trump_electoral_votes <- sum(prediction_comparison$electoral_votes[prediction_comparison$winner == "Trump"], na.rm = TRUE)

# Print the results
print(paste("Harris Electoral Votes:", harris_electoral_votes))
print(paste("Trump Electoral Votes:", trump_electoral_votes))

# Determine the predicted winner
if (harris_electoral_votes >= 270) {
  print("Kamala Harris is predicted to win the 2024 election.")
} else if (trump_electoral_votes >= 270) {
  print("Donald Trump is predicted to win the 2024 election.")
} else {
  print("No candidate reached 270 electoral votes.")
}

According to our models, Kamala Harris is projected to receive 216 electoral votes, while Donald Trump is predicted to win 147 electoral votes.

# Discussion {#sec-discuss}

## Key Findings

The multiple linear regression (MLR) models developed in this paper provide valuable insights into the factors that influence voter support for Kamala Harris and Donald Trump.
Through the analysis of polling data, several key predictors were identified as significant, including state, pollster rating (pollscore), and transparency score.

The MLR model for Kamala Harris highlights the importance of state-level factors in determining her support.
The model shows that voter preferences vary significantly across different states, with certain states (e.g., California and Maryland) showing higher support levels, while others (e.g., Indiana and Missouri) show lower support.
Pollster reliability, captured through numeric grades and transparency scores, also plays an important role, suggesting that voters may respond differently based on the credibility of the pollster.

Donald Trump Similarly, the MLR model for Donald Trump reveals that state-level variations are critical to understanding his level of support.
The model indicates that Trump’s support is more stable across certain states, but there are also notable outliers where his support fluctuates.
Pollster characteristics, such as pollscore and transparency score, significantly impact Trump’s predicted support, reflecting the importance of poll quality in predicting election outcomes.

Electoral College Forecast Aggregating state-level predictions into Electoral College votes demonstrates the potential competitiveness of the 2024 U.S.
Presidential Election.
Harris is projected to receive 216 electoral votes, while Trump is predicted to receive 147 electoral votes.
This forecast suggests that the election remains highly uncertain, with several key battleground states likely determining the final outcome.

@@

One of the main strengths of this analysis is the incorporation of multiple predictors that allow us to account for various factors influencing voter support.
By considering state-level data, pollster ratings, and transparency scores, the models provide a more nuanced prediction of voter behavior compared to models that rely solely on national-level polling.

The use of stepwise model selection and multicollinearity checks further enhanced the robustness of the models by optimizing the choice of predictors and ensuring that the models do not suffer from unstable estimates caused by correlated predictors.

Additionally, the aggregation of state-level predictions into Electoral College outcomes presents a more realistic forecast of the election, as the U.S.
Presidential election is ultimately decided by electoral votes, not popular votes.

@@

Despite the strengths of the models, there are several limitations that should be addressed:

- **Poll Reliability and Sampling Bias:** The accuracy of the predictions depends heavily on the quality of the polling data.
Although the model accounts for pollster ratings, polling methodologies can still introduce biases, particularly in states with limited polling data.
Sampling errors and non-response biases could skew the results, especially in smaller states or regions with inconsistent polling coverage.
- **Static Prediction:** The model provides a static snapshot of voter support based on current polling data, which may not capture the dynamic nature of voter preferences over time. As election day approaches, voter preferences may shift due to campaign events, debates, or other external factors.
Without time-series data, the model may fail to account for these trends.
- **Unaccounted Variables:** Although the model includes important predictors like state, pollscore, and transparency score, other potentially influential factors, such as economic conditions, campaign spending, and voter turnout, are not included in the analysis.These unaccounted variables may introduce inaccuracies in the final predictions.

## Future Research 
Directions Future work could improve upon this analysis by addressing some of the limitations mentioned above. For instance, incorporating time-series data could allow the model to capture how voter preferences evolve in response to external factors such as campaign events, economic developments, and political endorsements.
A dynamic forecasting model that updates predictions as new polls are released would provide more timely and accurate forecasts.

Moreover, integrating other influential variables, such as economic indicators (e.g., unemployment rates, inflation), voter turnout models, and campaign spending data, could enhance the predictive power of the model.
Including demographic data (e.g., age, education, income) could also improve the granularity of predictions, especially in battleground states where demographic shifts are critical to election outcomes.

Finally, expanding the model to account for ranked-choice voting in certain states could provide a more accurate forecast in scenarios where third-party candidates or run-off elections play a significant role.

@@

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-mlr-harris
#| fig-cap: Multi-Linear Regression model for Kamala Harris
#| fig.width: 10
#| fig.height: 4

mlr_harris_model <- readRDS(here::here(("models/mlr_harris_model.rds")))

# For Harris model
par(mfrow = c(1, 2))  # Set layout for 1 row and 2 columns
plot(mlr_harris_model, which = c(1, 2))  # 1: Residuals vs Fitted, 2: Normal Q-Q
par(mfrow = c(1, 1))  # Reset plot layout

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-mlr-trump
#| fig-cap: Multi-Linear Regression model for Donald Trump
#| fig.width: 10
#| fig.height: 4

# Load the saved MLR model for Donald Trump
mlr_trump_model <- readRDS(here::here(("models/mlr_trump_model.rds")))

# For Trump model
par(mfrow = c(1, 2))  # Set layout for 1 row and 2 columns
plot(mlr_trump_model, which = c(1, 2))  # 1: Residuals vs Fitted, 2: Normal Q-Q
par(mfrow = c(1, 1))  # Reset plot layout

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-harris-vif
#| fig-cap: Harris Refined Multi-Linear Regression Model
#| fig.width: 10
#| fig.height: 4

# Refine the model by removing less significant predictors 
# (e.g., methodology and transparency_score)
mlr_harris_model_refined <- readRDS(here::here(("models/mlr_harris_model_refined.rds")))

# For Harris refined model
par(mfrow = c(1, 2))  # Set layout for 1 row and 2 columns
plot(mlr_harris_model_refined, which = c(1, 2))  # 1: Residuals vs Fitted, 2: Normal Q-Q
par(mfrow = c(1, 1))  # Reset plot layout

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-harris-step
#| fig-cap: Harris Final Multi-Linear Regression Model
#| fig.width: 10
#| fig.height: 4

# Perform stepwise selection to optimize the Harris model
step_model <- step(mlr_harris_model_refined, direction = "both", trace=0)

# For Harris stepwise model
par(mfrow = c(1, 2))  # Set layout for 1 row and 2 columns
plot(step_model, which = c(1, 2))  # 1: Residuals vs Fitted, 2: Normal Q-Q
par(mfrow = c(1, 1))  # Reset plot layout

>>>>

---
title: "Forecasting the 2024 U.S. Presidential Election"
subtitle: "My subtitle if needed"
author: 
  - Tim Chen
  - Steven Li
  - Tommy Fu
thanks: "Code and data are available at: https://github.com/timchen0326/US_presidential_election_forecast_2024.git"
date: today
date-format: long
abstract: "Do Abstract after finalizing everything"
format: pdf
number-sections: true
toc: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

@@

Recent history has highlighted the complexities of election forecasting. The polling failures in 2016 and 2020—where polls significantly underestimated Republican support in key states—have prompted a fundamental reassessment of polling methodologies [@keeter2024]. These challenges are particularly acute in swing states, where margins of victory are often razor-thin and can determine the outcome of the entire election. The American Association for Public Opinion Research (AAPOR) identified several critical factors contributing to these polling errors, including the under representation of Republican voters and difficulties in predicting voter turnout patterns [@viala2024].

@@

This paper develops statistical models to forecast the outcome of the 2024 presidential election between Kamala Harris and Donald Trump. By leveraging multi-linear regression models, we predict the percentage of support for each candidate across different states, incorporating key variables such as pollster rating, sample size, and state-level demographics. Through aggregating these state-level predictions, we simulate Electoral College outcomes to provide insights into each candidate's probability of securing the required 270 electoral votes.Our analysis also includes a detailed examination of YouGov's polling methodology and proposes an idealized survey approach that could enhance the accuracy of election forecasting.

**Edit note: will need to expand/edit after getting final results**

@@

**Steven will do data section**

@@

- **Pollster rating (Pollscore):** A numerical score reflecting the reliability and historical accuracy of each polling organization.
- **Sample size:** The number of respondents included in each poll, which influences the poll's margin of error.
- **Support percentage (pct):** The percentage of respondents expressing support for each candidate.
- **State:** The U.S. state where the poll was conducted, or in some cases, national-level polling data.
- ...other variables

@tbl-overview_data in the appendix presents a sample of the processed dataset, showcasing the key variables necessary for the analysis.

@@

@@

# Model {#sec-model}

## Model Overview and Theoretical Framework

- Justify the choice of multiple linear regression for election forecasting
- Discuss why this approach is appropriate given the data structure
- Explain how the model addresses the unique challenges of electoral prediction
- Define the mathematical notation for the model
- Explain each component of the model specification
- Discuss underlying assumptions and their implications

## Baseline Model Development

- Present the initial model specification with all variables
- Explain the 80-20 train-test split methodology
- Detail the model validation approach
- Present mathematical notation:

Report key metrics such as, but not all required. U decide:

- R-squared values
- Mean Squared Error (MSE)
- Root Mean Squared Error (RMSE)
- Out-of-sample testing accuracy

## Model Refinement Using VIF Analysis

- Explain the importance of addressing multicollinearity
- Present VIF analysis methodology and thresholds used
- Show which variables were retained/removed and why
- Compare refined model performance to baseline
- Present updated mathematical notation
- Report comparative metrics showing improvement

@@

- Explain the stepwise selection methodology
- Justify the selection criteria (AIC/BIC)
- Present the final model specification
- Compare performance metrics across all three models
- Discuss why this represents the optimal model

## Model Validation and Diagnostics

- Present residual analysis
- Discuss normality assumptions
- Address heteroscedasticity
- Present cross-validation results
- Discuss model limitations

# Results {#sec-result}
## Model Performance Comparison

- Present comprehensive comparison table of all three models
- Include all relevant statistical measures
- Visualize performance metrics
- Discuss which model performs best and why

## State-Level Predictions

- Present predictions for key battleground states
- Compare model predictions with recent polling averages
- Discuss uncertainty in predictions
- Present confidence intervals

## Electoral College Projections

- Aggregate state-level predictions into electoral votes
- Present probability distributions for different outcomes
- Discuss implications for election outcome
- Address uncertainty in projections

## Model Insights

- Discuss which variables are most influential
- Present effect sizes for key predictors
- Analyze regional variations in model performance
- Identify any surprising or counterintuitive findings

# Discussion {#sec-discuss}
## Key Findings

- Summarize main results and their implications
- Compare findings with existing literature
- Discuss how results align with or differ from previous election modeling efforts
- Address practical significance of findings

@@

- Discuss advantages of the final model
- Compare with other approaches in literature
- Highlight unique contributions
- Explain practical applications

@@

- Address data limitations
- Discuss model assumptions and their impact
- Consider external validity
- Acknowledge potential sources of error

## Future Research Directions

- Suggest methodological improvements
- Propose additional variables to consider
- Discuss potential extensions of the model
- Recommend practical applications

@@

## Sample Data
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-overview_data
#| tbl-cap: Sample Overview of Selected Variables in the Polling Dataset

kable(t(data_overview))
```

@@
